`SCPCL000824` - Exploration of gene set based methods for tumor cell assignment

### 2024-07-03 analyses/cell-type-ewings/exploratory_analysis/annotation_notebooks/SCPCS000492/SCPCL000824_tumor-gene-expression-exploration.Rmd

```{r}
# add new column with updated marker gene classification 
# use local minima from 822 to define 824
new_classification <- combined_markers_df |> 
  dplyr::filter(sample == "SCPCL000824") |> 
  dplyr::mutate(updated_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
  dplyr::select(barcodes, updated_marker_gene_classification)

@@

For each gene set, several possible thresholds are calculated. 
The dotted lines over the histogram indicate the distribution and the corresponding vertical dotted line (same color) indicates the possible threshold. 
The thicker line indicates the chosen threshold by the algorithm. 
This corresponds to the highest value that reduces the false positives.

@@

```{r}
# compare AUCell classification vs cnv classification in ref
# pull out tumor cells found by AUCell
tumor_cells <- ref_assignment$markers$assignment

@@

# filter out any ambiguous cells 
caret_df <- ref_classification_df |> 
  dplyr::filter(tumor_cell_classification != "Ambiguous")

@@

It looks like we get pretty good agreement between our original classification and the updated `AUCell` classification using the marker gene list.

@@

all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df))

ggplot(all_auc_df, aes(x = auc, color = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_set))
```

Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values and tend to mirror each other. 
The main difference here is the lack of bimodal distribution in `SCPCL000824`.

Let's use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
Here, we label any cells that have passed the threshold set for the marker gene set as tumor cells.

```{r}
# pull out threshold used for assigning cells from ref sample 
ref_threshold <- ref_assignment$markers$aucThr$selected

@@

This looks similar to using the updated marker gene cutoff to classify cells where most cells in that bottom group are tumor cells.  
As a reminder, the updated marker gene classification is the classification of tumor cells based on the marker gene expression cutoff determined in `SCPCL000822`. 
We can confirm this by calculating the confusion matrix.

```{r}
# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$updated_marker_gene_classification, 
      classification_df$auc_classification)
  ) 
```

@@

# add classification column for comparison to original classification
caret_df <- caret_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal")
  )

@@

It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
Now we will use the marker gene set cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.

@@

# add ucell classification
classification_df <- classification_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal")
  )

@@

This looks similar to what we get with both the updated marker gene and `AUCell` classification. 
We can confirm that by calculating the confusion matrix.

>>>>

```{r}
# add new column with updated marker gene classification 
# use local minima from 822 to define 824
new_classification <- combined_markers_df |> 
  dplyr::filter(sample == "SCPCL000824") |> 
  dplyr::mutate(updated_marker_gene_classification = dplyr::if_else(sum_raw_exp >= exp_cutoff, "Tumor", "Normal")) |> 
  dplyr::select(barcodes, updated_marker_gene_classification) |> 
  unique()

@@

For each gene set, several possible thresholds are calculated. 
The dotted lines over the histogram indicate the distribution and the corresponding vertical dotted line (same color) indicates the possible threshold. 
The thicker line indicates the chosen threshold by the algorithm. 
This corresponds to the highest value that reduces the false positives. 
See the [`AUCell` vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/AUCell/inst/doc/AUCell.html#determine-the-cells-with-the-given-gene-signatures-or-active-gene-sets) for more information.

@@

```{r}
# compare AUCell classification vs cnv classification in ref
# pull out tumor cells found by AUCell
tumor_cells <- ref_assignments$markers$assignment

@@

# filter out any ambiguous cells 
caret_df <- ref_classification_df |> 
  dplyr::filter(tumor_cell_classification != "Ambiguous") |> 
  # make sure positive class is tumor
  dplyr::mutate(
    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
  )

@@

Here the original tumor cell classifications are the rows and the `AUCell` classification are the columns. 
It looks like we get pretty good agreement between our original classification and the updated `AUCell` classification using the marker gene list.

@@

all_auc_df <- dplyr::bind_rows(list(ref_auc_df, new_auc_df)) |> 
  # simplify the name for plotting
  dplyr::mutate(gene_set = stringr::word(gene_set, 1, sep = "_"))

ggplot(all_auc_df, aes(x = auc, color = sample)) +
  geom_density() +
  facet_grid(rows = vars(gene_set),
             scales = "free_y")
```

Looking at the AUC across both samples for all gene sets, we see that they are in the same range of values. 
The main difference here is the lack of bimodal distribution in `SCPCL000824` and we see that most of the time the peak for `SCPCL000824` seems to be in between the two peaks for `SCPCL000822`.

The exception to this is with the marker gene list where the peak in `SCPCL000824` lines up with the second peak in `SCPCL000822`. 
Because of that we will use the marker gene threshold identified by `AUCell` for `SCPCL000822` to identify tumor cells in `SCPCL000824`.  
Here, we label any cells that have passed the threshold set for the marker gene set as tumor cells.

```{r}
# pull out threshold used for assigning cells from ref sample 
ref_threshold <- ref_assignments$markers$aucThr$selected

@@

This looks similar to using the updated marker gene cutoff to classify cells where most cells in that bottom group are tumor cells.  
As a reminder, the updated marker gene classification is the classification of tumor cells based on the marker gene expression cutoff determined in `SCPCL000822`. 
We can confirm this by calculating the confusion matrix. 
Here the marker gene annotations are the rows and the annotations from `AUCell` are the columns.

```{r}
classification_df <- classification_df |> 
  dplyr::mutate(
    updated_marker_gene_classification = forcats::fct_relevel(updated_marker_gene_classification, "Tumor"),
    auc_classification = forcats::fct_relevel(auc_classification, "Tumor")
  )

# compare using a confusion matrix
caret::confusionMatrix(
    table(
      classification_df$updated_marker_gene_classification, 
      classification_df$auc_classification)
  ) 
```

@@

# add classification column for comparison to original classification
caret_df <- caret_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% ref_ucell_tumor_cells, "Tumor", "Normal"), 
    tumor_cell_classification = forcats::fct_relevel(tumor_cell_classification, "Tumor"),
    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
  )

@@

Here the rows correspond to the original tumor annotations and the columns are the `UCell` annotations. 
It looks like we get pretty good agreement between using `UCell` with the `RIGGI` gene set to the original classification. 
Now we will use the marker gene set cutoff from `SCPCL000822` to classify tumor cells in `SCPCL000824`.

@@

# add ucell classification
classification_df <- classification_df |> 
  dplyr::mutate(
    ucell_classification = dplyr::if_else(barcodes %in% new_ucell_tumor_cells, "Tumor", "Normal"), 
    # relevel for confusion matrix later 
    updated_marker_gene_classification = forcats::fct_relevel(updated_marker_gene_classification, "Tumor"),
    ucell_classification = forcats::fct_relevel(ucell_classification, "Tumor")
  )

@@

This looks similar to what we get with both the updated marker gene and `AUCell` classification. 
We can confirm that by calculating the confusion matrix.
Here the rows correspond to the updated marker gene annotations and the columns are the `UCell` annotations.
